Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-10-10 paper/paper_draft.qmd
Add LTMP citation to figure caption
Update citation formatting

Findings on the importance of connectivity likely differ from literature due to the highly
uncertain nature of larval connectivity modelling [@choukroun2025; @vogt-vincent2023], the
way larval flows and juvenile recruitment are implemented in ADRIA - CoralBlox, and the
magnitude of early bleaching events decreasing the breeding capacity of strongly connected
reefs. Analysing daily connectivity data Vogt-Vincent et al. [-@vogt-vincent2023] suggest
that time-mean connectivity may not be suitable for capturing the seasonal - ecological time
scale connectivity between locations due to high daily variability in connectivity data.
Connectivity matrices used in this investigation have large relative standard deviations for
larval transition probabilities between reefs over the 24 days used
(@sfig-connectivity-rsd). Using alternative summarisation methods that capture
ecological-timescale patterns in coral larval dispersal, such as cumulated implicit
connectivity [@vogt-vincent2023], future investigations may find increased importance of
both depth and reef connectivity for resilience, in line with other investigations
[@hock2017; @torres2024]. Choukroun et al [-@choukroun2025] performed a comparison of three
GBR-scale connectivity models, including GBR1, and suggest the use of a model ensemble
approach to estimate larval dispersal due to the high sensitivity of connectivity
predictions to model specification. Further improvements and research in connectivity
modelling are necessary when incorporating connectivity as a supplementary factor in reef
resilience and intervention-planning analyses.

@@

In addition to the usual challenges of deriving data from satellite imagery, it is
acknowledged that there is high potential for misalignment between the bathymetry data and
GBRMPA reef outlines. Hammerton and Lawrey [-@hammerton2025], for example, identified 120
false positives, where features identified as reefs were sand banks or rocky reefs (rather
than coral reefs). Reefs with bathymetry data covering less than 5% of their indicative area
were excluded from analyses (see @sfig-depth-cutoff). Existing GBRMPA reef outlines were
adopted to align with other ongoing modelling and analyses conducted as part of the Reef
Restoration and Adaptation Program (<https://gbrrestoration.org>; [@thereef]).

@@

Monthly mean SST climatological baselines, for each 1km^2^ pixel, were calculated from the
daily SST data following NOAA methodology [@liu2017; @skirving2020]. For each pixel of SST
data, the Monthly Mean SST climatology was calculated by computing the mean SST value for
each calendar month (January through December) across the 28-year baseline period from 1985
to 2012. For each month separately, this climatological baseline was adjusted to centre at
year 1988.2857, which attempts to account for any long-term trends within the baseline years
themselves (see Skirving et al. [@skirving2020]). The adjustment ensures that the
climatology represents conditions at a specific temporal reference point rather than being
influenced by gradual warming or cooling trends that may have occurred during the 1985-2012
period. The resulting monthly climatologies serve as a standardized reference against which
current SST conditions can be compared to identify thermal anomalies and assess coral
bleaching risk, following the methodology established by Skirving et al. [-@skirving2020].

@@

![Time series comparing modelled and observed relative coral cover for Ben reef showing a
recovery process under very low disturbance and no interventions. The observed data used
here is from the AIMS Long-Term Monitoring Program (LTMP). Error metrics including Root Mean
Square Error (0.06) and the Spearman Rank Correlation Coefficient (0.9) is also shown. The
second plot shows the DHW and Cyclone or COTS events. The third plot shows the benthic
composition for each functional group as a proportion of the reef’s cover at each
year.](../figs/coralblox_reef_example.png){#sfig-coralblox-example width fig-align="center"
width="14.82cm"}

@@

![Parametric sensitivity analysis conducted with Shapley effects. Each column reflects
sampling of different DHW ranges. The first column represents model runs with DHW = 0, and
the remaining columns represent models runs with DHW valued between 0 and 8, 8 and 16, 16
and 30, and 0 and 30. The first row shows the Shapley effect for change in coral cover and
the second for change in evenness of species
abundance.](../figs/coralblox_sens_analysis.png){#sfig-coralblox-sa width fig-align="center"
width="14.82cm"}

>>>>

Findings on the importance of connectivity likely differ from literature due to the highly
uncertain nature of larval connectivity modelling [@choukroun2025; @vogt-vincent2023], the
way larval flows and juvenile recruitment are implemented in ADRIA - CoralBlox, and the
magnitude of early bleaching events decreasing the breeding capacity of strongly connected
reefs. Analysing daily connectivity data Vogt-Vincent, Mitarai and Johnson [-@vogt-vincent2023] suggest
that time-mean connectivity may not be suitable for capturing the seasonal - ecological time
scale connectivity between locations due to high daily variability in connectivity data.
Connectivity matrices used in this investigation have large relative standard deviations for
larval transition probabilities between reefs over the 24 days used
(@sfig-connectivity-rsd). Using alternative summarisation methods that capture
ecological-timescale patterns in coral larval dispersal, such as cumulated implicit
connectivity [@vogt-vincent2023], future investigations may find increased importance of
both depth and reef connectivity for resilience, in line with other investigations
[@hock2017; @torres2024]. Choukroun et al [-@choukroun2025] performed a comparison of three
GBR-scale connectivity models, including GBR1, and suggest the use of a model ensemble
approach to estimate larval dispersal due to the high sensitivity of connectivity
predictions to model specification. Further improvements and research in connectivity
modelling are necessary when incorporating connectivity as a supplementary factor in reef
resilience and intervention-planning analyses.

@@

In addition to the usual challenges of deriving data from satellite imagery, it is
acknowledged that there is high potential for misalignment between the bathymetry data and
GBRMPA reef outlines. Hammerton and Lawrey [-@hammerton2025], for example, identified 120
false positives, where features identified as reefs were sand banks or rocky reefs (rather
than coral reefs). Reefs with bathymetry data covering less than 5% of their indicative area
were excluded from analyses (see @sfig-depth-cutoff). Existing GBRMPA reef outlines were
adopted to align with other ongoing modelling and analyses conducted as part of the Reef
Restoration and Adaptation Program (<https://gbrrestoration.org>; RRAP [-@thereef]).

@@

Monthly mean SST climatological baselines, for each 1km^2^ pixel, were calculated from the
daily SST data following NOAA methodology [@liu2017; @skirving2020]. For each pixel of SST
data, the Monthly Mean SST climatology was calculated by computing the mean SST value for
each calendar month (January through December) across the 28-year baseline period from 1985
to 2012. For each month separately, this climatological baseline was adjusted to centre at
year 1988.2857, which attempts to account for any long-term trends within the baseline years
themselves (see Skirving et al. [-@skirving2020]). The adjustment ensures that the
climatology represents conditions at a specific temporal reference point rather than being
influenced by gradual warming or cooling trends that may have occurred during the 1985-2012
period. The resulting monthly climatologies serve as a standardized reference against which
current SST conditions can be compared to identify thermal anomalies and assess coral
bleaching risk, following the methodology established by Skirving et al. [-@skirving2020].

@@

![Time series comparing modelled and observed relative coral cover for Ben reef showing a
recovery process under very low disturbance and no interventions. The observed data used
here is from the AIMS Long-Term Monitoring Program (LTMP) . Error metrics including Root
Mean Square Error (0.06) and the Spearman Rank Correlation Coefficient (0.9) is also shown.
The second plot shows the DHW and Cyclone or COTS events. The third plot shows the benthic
composition for each functional group as a proportion of the reef’s cover at each
year.](../figs/coralblox_reef_example.png){#sfig-coralblox-example
fig-align="center" width="14.82cm"}

@@

![Parametric sensitivity analysis conducted with Shapley effects. Each column reflects
sampling of different DHW ranges. The first column represents model runs with DHW = 0, and
the remaining columns represent models runs with DHW valued between 0 and 8, 8 and 16, 16
and 30, and 0 and 30. The first row shows the Shapley effect for change in coral cover and
the second for change in evenness of species
abundance.](../figs/coralblox_sens_analysis.png){#sfig-coralblox-sa
fig-align="center" width="14.82cm"}
